%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DS9 Integrated Source Extractor
% Technical Reference Manual
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%    PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt,fleqn]{book}

\usepackage[top=3cm,bottom=3cm,left=3.2cm,right=3.2cm,headsep=10pt,headheight=14pt,a4paper]{geometry}

\usepackage{xcolor}
\definecolor{ocre}{RGB}{52,177,201}

% Font Settings
\usepackage{avant}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

% Bibliography
\usepackage[style=alphabetic,sorting=nyt,sortcites=true,autopunct=true,hyperref=true,abbreviate=false,backref=true,backend=biber]{biblatex}
\addbibresource{bibliography.bib}
\defbibheading{bibempty}{}

\input{structure}

%----------------------------------------------------------------------------------------
%    Definitions of new commands
%----------------------------------------------------------------------------------------

\def\R{\mathbb{R}}
\newcommand{\ds}{\texttt{ds9}}
\newcommand{\sextract}{\texttt{ds9\_sextract}}
\newcommand{\sep}{\texttt{SEP}}

%------------------------------------------------------
% Document
%------------------------------------------------------
\begin{document}

%------------------------------------------------------
% Title page
%------------------------------------------------------
\begin{titlepage}
\begin{tikzpicture}[remember picture,overlay]
  % Top colored area
  \fill[ocre] (current page.north west) rectangle ([yshift=-9cm]current page.north east);
  % Bottom accent strip
  \fill[ocre!10] (current page.south west) rectangle ([yshift=2.5cm]current page.south east);
  \draw[ocre,line width=1.5pt] ([yshift=2.5cm]current page.south west) -- ([yshift=2.5cm]current page.south east);
\end{tikzpicture}

\vspace*{1cm}

\begin{center}
{\fontsize{38}{46}\selectfont\sffamily\bfseries\color{white}DS9 Integrated}\\[0.5cm]
{\fontsize{38}{46}\selectfont\sffamily\bfseries\color{white}Source Extractor}\\[1cm]
{\fontsize{20}{24}\selectfont\sffamily\color{white!80}Technical Reference Manual}
\end{center}

\vspace{5cm}

\begin{center}
{\Large\sffamily
SEP-Based Source Detection, Photometry,\\[8pt]
and Catalog Management\\[8pt]
Integrated into SAOImageDS9}

\vspace{3cm}

{\large\sffamily Juhan~Kim}\\[0.5cm]
{\sffamily March 2026}
\end{center}

\vfill

\begin{center}
{\small\sffamily
Based on SAOImageDS9 (Smithsonian Astrophysical Observatory)\\[3pt]
and SEP --- Source Extraction and Photometry (Barbary et al.)}
\end{center}
\vspace{1cm}
\end{titlepage}

%------------------------------------------------------
% Table of Contents
%------------------------------------------------------

\pagestyle{empty}
\tableofcontents
\pagestyle{fancy}

%------------------------------------------------------
% PART I
%------------------------------------------------------
\part{User Guide}

%======================================================
\chapter{Quick Start Guide}
\label{ch:quickstart}
%======================================================

This chapter provides a step-by-step guide to using the integrated Source Extractor in DS9. For detailed explanations of the extraction algorithm, output parameters, and internal architecture, refer to the subsequent chapters.

\section{Overview}

The DS9 Integrated Source Extractor adds SExtractor-like source detection and photometry capabilities directly within the SAOImageDS9 image viewer. When a FITS image is loaded, sources can be extracted with a single click, and the resulting catalog is displayed in a resizable panel alongside the image.

Key features:
\begin{itemize}[leftmargin=2em]
  \item \textbf{One-click extraction} --- Press the ``Extract'' button to detect and measure all sources in the current FITS image
  \item \textbf{Split-panel interface} --- The main window is divided into an image viewer (left) and a catalog table (right), separated by a draggable divider
  \item \textbf{19 output parameters} --- Photometry, astrometry, morphology, surface brightness, and star/galaxy classification
  \item \textbf{Real-time filtering} --- Filter the catalog interactively by any column value
  \item \textbf{Automatic WCS} --- Celestial coordinates (RA, Dec) are computed automatically when WCS information is available in the FITS header
  \item \textbf{Auto-extraction} --- Sources are extracted automatically when a new FITS file is loaded
\end{itemize}

\section{Prerequisites}

\begin{itemize}[leftmargin=2em]
  \item \textbf{DS9} --- SAOImageDS9 built with the integrated Source Extractor patch
  \item \textbf{Python 3} with the following packages:
  \begin{itemize}
    \item \texttt{sep} --- Source Extraction and Photometry library (Python binding of SExtractor)
    \item \texttt{astropy} --- FITS I/O and WCS coordinate transformations
    \item \texttt{numpy} --- Numerical array operations
  \end{itemize}
  \item \textbf{\sextract{} binary} --- The compiled Python script, located in the same directory as the \ds{} executable
\end{itemize}

\begin{codebox}[Install Python Dependencies]
\begin{lstlisting}[style=bash,numbers=none]
pip install sep astropy numpy
\end{lstlisting}
\end{codebox}

\section{Building DS9 with the Source Extractor}

\begin{codebox}[Build Commands]
\begin{lstlisting}[style=bash,numbers=none]
cd SAOImageDS9
make -C ds9/unix -j 128
\end{lstlisting}
\end{codebox}

The build process compiles the C/Tcl core, packages the Tcl library files (including the modified \texttt{layout.tcl} and \texttt{ds9.tcl}) into \texttt{ds9.zip}, and produces the final \texttt{bin/ds9} executable. The \sextract{} binary must be placed in the same \texttt{bin/} directory.

\begin{important}
After modifying any \texttt{.tcl} file in \texttt{ds9/library/}, you must copy it to \texttt{ds9/unix/mntpt/library/}, update the zip, and rebuild:
\begin{codebox}
\begin{lstlisting}[style=bash,numbers=none]
cp ds9/library/layout.tcl ds9/unix/mntpt/library/
cp ds9/library/ds9.tcl    ds9/unix/mntpt/library/
cd ds9/unix
zip -ru ds9.zip mntpt/library/layout.tcl mntpt/library/ds9.tcl
cat ds9base ds9.zip > ds9
zip -A ds9
chmod 755 ds9
cp ds9 ../../bin/
\end{lstlisting}
\end{codebox}
\end{important}

\section{Using the Source Extractor}

\subsection{Step 1: Launch DS9}

\begin{codebox}[Launch]
\begin{lstlisting}[style=bash,numbers=none]
./bin/ds9 image.fits
\end{lstlisting}
\end{codebox}

The window opens with two panels separated by a vertical divider:
\begin{itemize}[leftmargin=2em]
  \item \textbf{Left panel} --- The standard DS9 interface (menu bar, info panel, buttons, image canvas, colorbar)
  \item \textbf{Right panel} --- The Source Extractor catalog panel (title bar, filter bar, data table, status bar)
\end{itemize}

The divider can be dragged left or right to resize the panels.

\subsection{Step 2: Extract Sources}

Click the \textbf{``Extract''} button in the catalog panel title bar. The status bar will show progress:
\begin{enumerate}[leftmargin=2em]
  \item \texttt{Extracting sources from image.fits ...} --- Extraction is running
  \item \texttt{image.fits: 142 sources extracted} --- Extraction complete
\end{enumerate}

The catalog table is automatically populated with all detected sources, sorted by brightness (brightest first).

\begin{infobox}[Auto-Extraction]
When a FITS file is loaded via \texttt{File $\to$ Open}, sources are extracted automatically after a 500\,ms delay. This behavior is controlled by the \texttt{CatalogPanelAutoExtract} hook in \texttt{fits.tcl}.
\end{infobox}

\subsection{Step 3: Browse the Catalog}

The catalog table displays 19 columns (see Chapter~\ref{ch:parameters} for detailed descriptions):

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Category} & \textbf{Columns} \\
\midrule
Identification & NUMBER \\
Astrometry & X\_IMAGE, Y\_IMAGE, ALPHA\_J2000, DELTA\_J2000 \\
Photometry & MAG\_AUTO, MAG\_ISOCOR, MAG\_APER, FLUX\_AUTO \\
Morphology & FLUX\_RADIUS, A\_IMAGE, B\_IMAGE, THETA\_IMAGE, ELLIPTICITY \\
 & KRON\_RADIUS \\
Surface Brightness & MU\_MAX, MU\_THRESHOLD \\
Classification & CLASS\_STAR, FLAGS \\
\bottomrule
\end{tabular}
\end{center}

\begin{itemize}[leftmargin=2em]
  \item Scroll horizontally and vertically using the scrollbars
  \item Column widths can be resized by dragging column borders in the header row
  \item Rows can be selected by clicking (extended selection with Shift/Ctrl)
\end{itemize}

\subsection{Step 4: Filter Results}

Type a search pattern in the \textbf{Filter} entry and click \textbf{``Apply''} (or press Enter). The filter performs case-insensitive substring matching across all columns. For example:
\begin{itemize}[leftmargin=2em]
  \item Type \texttt{0.9} to show only rows containing ``0.9'' in any field
  \item Clear the filter field and press Apply to show all sources again
\end{itemize}

\subsection{Step 5: Clear and Re-extract}

Click \textbf{``Clear''} to reset the table and status. Load a new FITS image and click \textbf{``Extract''} again, or rely on auto-extraction.


\section{Interface Layout}

The window hierarchy below the pulldown menu bar:

\begin{codebox}[Widget Hierarchy]
\begin{lstlisting}[numbers=none]
. (top-level window)
|-- .mb (pulldown menu bar: File, Edit, View, ...)
|-- .pw (ttk::panedwindow, horizontal, draggable divider)
    |-- .ds9 (left pane: main DS9 interface)
    |   |-- header (info panel, magnifier, panner)
    |   |-- buttons (toolbar)
    |   |-- image frame
    |       |-- canvas (FITS image display)
    |       |-- colorbar, graphs
    |-- .catf (right pane: catalog panel)
        |-- titlebar ("Source Extractor", Extract, Clear)
        |-- searchbar (Filter entry + Apply)
        |-- table (19-column scrollable data table)
        |-- statusbar (status messages)
\end{lstlisting}
\end{codebox}

The \texttt{ttk::panedwindow} widget provides a sash (divider handle) that the user can drag to redistribute space between the image viewer and the catalog panel. The left pane has weight~3 and the right pane has weight~2, giving an initial 60:40 split.

\section{Troubleshooting}

\begin{center}
\begin{longtable}{L{4.5cm} L{8.5cm}}
\toprule
\textbf{Symptom} & \textbf{Solution} \\
\midrule
\endhead
\texttt{ds9\_sextract not found} & Ensure the \sextract{} binary is in the same directory as the \ds{} executable. Check that it has execute permission (\texttt{chmod +x}). \\
\midrule
\texttt{No FITS image loaded} & Open a FITS file before clicking Extract. The extractor requires a frame with a loaded FITS image. \\
\midrule
\texttt{Extraction error: ...} & Check that Python dependencies (\texttt{sep}, \texttt{astropy}, \texttt{numpy}) are installed. Check \texttt{LD\_LIBRARY\_PATH} if using conda. \\
\midrule
\texttt{No sources detected} & The image may have very low signal-to-noise. Try adjusting the detection threshold (default: $1.5\sigma$) via command-line arguments to \sextract{}. \\
\midrule
Catalog panel not visible & Drag the divider from the right edge of the window toward the center. The panel may have been collapsed. \\
\midrule
WCS coordinates show 0.0 & The FITS header does not contain valid WCS keywords. Pixel coordinates (X\_IMAGE, Y\_IMAGE) are still accurate. \\
\bottomrule
\end{longtable}
\end{center}


%======================================================
\chapter{Architecture}
\label{ch:architecture}
%======================================================

This chapter describes the internal architecture of the integrated Source Extractor, covering the modified DS9 source files, the extraction pipeline, and the data flow between components.

\section{Modified Source Files}

Three files implement the integration:

\begin{center}
\begin{longtable}{L{4.2cm} L{8.8cm}}
\toprule
\textbf{File} & \textbf{Role} \\
\midrule
\endhead
\texttt{ds9/library/ds9.tcl} & Creates the top-level \texttt{ttk::panedwindow} that splits the main window into two panes. Initializes \texttt{ds9(toppw)}, \texttt{ds9(main)}, and \texttt{ds9(catalog\_frame)}. Calls \texttt{CreateCatalogPanel}. \\
\midrule
\texttt{ds9/library/layout.tcl} & Defines \texttt{CreateCanvas} (image canvas, restored to original structure), \texttt{CreateCatalogPanel} (catalog UI), \texttt{CatalogPanelExtract} (runs \sextract{}), \texttt{CatalogPanelLoadTSV} (parses output into table), \texttt{CatalogPanelFilter} (interactive filtering), and \texttt{CatalogPanelAutoExtract} (auto-extraction hook). \\
\midrule
\texttt{ds9/library/fits.tcl} & Calls \texttt{CatalogPanelAutoExtract} after a FITS file is loaded, triggering automatic source extraction. \\
\bottomrule
\end{longtable}
\end{center}

\section{Top-Level Window Structure}

The key architectural change is the introduction of a \texttt{ttk::panedwindow} at the highest level of the widget hierarchy, directly below the native pulldown menu bar.

\begin{codebox}[ds9.tcl --- Top-level panedwindow creation]
\begin{lstlisting}[style=tcl,numbers=none]
# create top-level panedwindow
set ds9(toppw) [ttk::panedwindow ${ds9(top)}pw -orient horizontal]
pack $ds9(toppw) -fill both -expand true

# left pane: main DS9 content
set ds9(main) [ttk::frame ${ds9(top)}ds9]
$ds9(toppw) add $ds9(main) -weight 3

# right pane: catalog panel
set ds9(catalog_frame) [ttk::frame ${ds9(top)}catf]
$ds9(toppw) add $ds9(catalog_frame) -weight 2
\end{lstlisting}
\end{codebox}

This design ensures that:
\begin{enumerate}[leftmargin=2em]
  \item The pulldown menu bar spans the full window width (it is set via \texttt{\$ds9(top) configure -menu \$ds9(mb)}, independent of pack geometry)
  \item Everything below the menu bar is split into two resizable panes
  \item All existing DS9 layout modes (horizontal, vertical, basic, advanced) continue to work within the left pane without modification
  \item The catalog panel occupies its own independent pane on the right
\end{enumerate}

\section{Data Flow}

The extraction pipeline follows this sequence:

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Trigger} --- User clicks ``Extract'' (or auto-extraction fires after FITS load)
  \item \textbf{File discovery} --- \texttt{CatalogPanelExtract} retrieves the current FITS filename from the active frame via \texttt{\$current(frame) get fits file name full}
  \item \textbf{External process} --- The \sextract{} binary is executed as a child process via \texttt{exec sh -c}. The \texttt{LD\_LIBRARY\_PATH} is configured to include conda library paths.
  \item \textbf{Output capture} --- \sextract{} writes a tab-separated table to \texttt{stdout}. The Tcl \texttt{exec} command captures this output as a string.
  \item \textbf{Table population} --- \texttt{CatalogPanelLoadTSV} parses the TSV string, splits it into rows and columns, and populates the \texttt{Tktable} widget via the \texttt{catpaneltbldb} array variable.
  \item \textbf{Status update} --- The status bar displays the filename and number of detected sources.
\end{enumerate}

\begin{remark}[Process Isolation]
The extraction runs as a separate process, not within the DS9 Tcl interpreter. This prevents numerical library conflicts and allows the Python-based extractor to use its own dependencies (NumPy, SEP, Astropy) independently.
\end{remark}


%======================================================
\chapter{Extraction Algorithm}
\label{ch:algorithm}
%======================================================

The \sextract{} tool implements a SExtractor-compatible extraction pipeline using the SEP (Source Extraction and Photometry) library. This chapter describes each step of the algorithm in detail.

\section{Background Estimation}

The first step is to estimate and subtract the sky background. SEP uses the same algorithm as SExtractor:

\begin{enumerate}[leftmargin=2em]
  \item The image is divided into a grid of rectangular meshes (default size: $64 \times 64$ pixels, controlled by \texttt{--back-size})
  \item In each mesh cell, the local background level is estimated using iterative $\kappa\sigma$-clipping
  \item The resulting coarse background map is smoothed with a median filter (default kernel: $3 \times 3$ cells, controlled by \texttt{--back-filtersize})
  \item The smoothed map is interpolated to full image resolution using bicubic spline interpolation
\end{enumerate}

\begin{codebox}[Background estimation]
\begin{lstlisting}[style=python,numbers=none]
bkg = sep.Background(data, bw=back_size, bh=back_size,
                     fw=back_filtersize, fh=back_filtersize)
data_sub = data - bkg.back()
\end{lstlisting}
\end{codebox}

Two outputs are produced:
\begin{itemize}[leftmargin=2em]
  \item \texttt{bkg.back()} --- the background level map $B(x,y)$
  \item \texttt{bkg.rms()} --- the background RMS (noise) map $\sigma_B(x,y)$, used as the detection threshold and photometric error
\end{itemize}

\section{Source Detection}

Sources are detected as groups of connected pixels above a threshold:

\begin{equation}
I(x,y) - B(x,y) > k \cdot \sigma_B(x,y)
\end{equation}

where $k$ is the detection threshold in units of the local background RMS (default: $k = 1.5$, set by \texttt{--detect-thresh}). Connected pixel groups with fewer than \texttt{--detect-minarea} pixels (default: 5) are rejected.

\subsection{Deblending}

Overlapping sources are separated using multi-threshold deblending:

\begin{enumerate}[leftmargin=2em]
  \item The detected object is re-thresholded at \texttt{--deblend-nthresh} (default: 32) exponentially spaced sub-thresholds between the detection threshold and the peak value
  \item At each sub-threshold, connected components are identified
  \item A component is split into a separate object if it contains more than \texttt{--deblend-mincont} (default: 0.005) of the total flux of the parent object
\end{enumerate}

\begin{codebox}[Source detection with deblending]
\begin{lstlisting}[style=python,numbers=none]
objects, segmap = sep.extract(
    data_sub, detect_thresh, err=bkg_rms,
    minarea=detect_minarea,
    deblend_nthresh=deblend_nthresh,
    deblend_cont=deblend_mincont,
    segmentation_map=True
)
\end{lstlisting}
\end{codebox}

The segmentation map assigns each pixel to the object it belongs to (or zero for background), enabling morphological measurements.


\section{Photometry}

Three photometric methods are applied to each detected source.

\subsection{AUTO Photometry (Kron Elliptical Aperture)}

The AUTO magnitude follows the Kron (1980) method:

\begin{enumerate}[leftmargin=2em]
  \item The first-moment radius (Kron radius $r_K$) is computed within an ellipse of semi-major axis $6a$:
  \begin{equation}
    r_K = \frac{\sum_i r_i \, I_i}{\sum_i I_i}
  \end{equation}
  where $r_i$ is the elliptical distance of pixel $i$ from the object center, and $I_i$ is the background-subtracted intensity.
  \item A minimum Kron radius of 3.5 pixels is enforced to prevent underestimation for compact sources.
  \item The flux is summed within an elliptical aperture of semi-major axis $2.5 \, r_K$, following the SExtractor convention.
\end{enumerate}

\begin{infobox}[Kron Factor]
The factor of 2.5 applied to the Kron radius ensures that ${\gtrsim}96\%$ of the total flux is captured for a typical galaxy profile (S\'{e}rsic $n \approx 1$--4), while ${\gtrsim}99\%$ is captured for a Gaussian PSF.
\end{infobox}

\subsection{Isophotal Corrected Photometry}

The isophotal flux is the sum of all pixels belonging to the detection footprint. A correction factor accounts for the flux missed below the detection isophote, using the SExtractor formula:

\begin{equation}
F_{\text{isocor}} = \frac{F_{\text{iso}}}{1 - 0.196\,\alpha - 0.751\,\alpha^2}, \qquad
\alpha = \frac{N_{\text{pix}} \cdot t_{\text{local}}}{F_{\text{iso}}}
\end{equation}

where $N_{\text{pix}}$ is the number of detection pixels, $t_{\text{local}}$ is the local threshold, and $F_{\text{iso}}$ is the raw isophotal flux.

\subsection{Aperture Photometry}

A fixed circular aperture of diameter \texttt{--phot-aperture} (default: 5.0 pixels) is centered on each source. The flux is summed within this circle, with partial pixel weighting at the boundary.

\subsection{Magnitude Conversion}

All fluxes are converted to magnitudes using:
\begin{equation}
m = -2.5 \log_{10}(F) + m_0
\end{equation}

where $m_0$ is the magnitude zero-point (default: 25.0, set by \texttt{--mag-zeropoint}). Sources with non-positive flux are assigned $m = 99.0$.


\section{Morphological Parameters}

\subsection{Shape Parameters}

The semi-major axis ($a$), semi-minor axis ($b$), and position angle ($\theta$) are derived from the second-order moments of the intensity distribution within the detection isophote:

\begin{align}
\overline{x^2} &= \frac{\sum_i I_i \, x_i^2}{\sum_i I_i} - \bar{x}^2 \\
\overline{y^2} &= \frac{\sum_i I_i \, y_i^2}{\sum_i I_i} - \bar{y}^2 \\
\overline{xy}  &= \frac{\sum_i I_i \, x_i \, y_i}{\sum_i I_i} - \bar{x}\bar{y}
\end{align}

The ellipse parameters are:
\begin{align}
a^2 &= \frac{\overline{x^2} + \overline{y^2}}{2} + \sqrt{\left(\frac{\overline{x^2} - \overline{y^2}}{2}\right)^2 + \overline{xy}^2} \\
b^2 &= \frac{\overline{x^2} + \overline{y^2}}{2} - \sqrt{\left(\frac{\overline{x^2} - \overline{y^2}}{2}\right)^2 + \overline{xy}^2} \\
\theta &= \frac{1}{2}\arctan\!\left(\frac{2\,\overline{xy}}{\overline{x^2} - \overline{y^2}}\right)
\end{align}

The ellipticity is defined as:
\begin{equation}
e = 1 - \frac{b}{a}
\end{equation}

\subsection{Half-Light Radius}

The half-light radius (\texttt{FLUX\_RADIUS}) is the radius of a circular aperture centered on the source that encloses 50\% of the AUTO flux. It is computed by integrating the flux profile outward from the center until the cumulative flux reaches $0.5 \times F_{\text{AUTO}}$.

\section{Surface Brightness}

Two surface brightness measurements are provided:

\begin{itemize}[leftmargin=2em]
  \item \textbf{MU\_MAX} --- Surface brightness of the brightest pixel:
  \begin{equation}
    \mu_{\max} = -2.5\log_{10}\!\left(\frac{I_{\text{peak}}}{\Omega_{\text{pix}}}\right) + m_0
  \end{equation}
  where $I_{\text{peak}}$ is the peak pixel value above background and $\Omega_{\text{pix}}$ is the pixel area in arcsec$^2$.

  \item \textbf{MU\_THRESHOLD} --- Surface brightness at the detection threshold:
  \begin{equation}
    \mu_{\text{thresh}} = -2.5\log_{10}\!\left(\frac{t_{\text{local}}}{\Omega_{\text{pix}}}\right) + m_0
  \end{equation}
\end{itemize}


\section{Star/Galaxy Classification}

An approximate star/galaxy classifier is implemented using the ratio of the measured FWHM to the seeing FWHM:

\begin{equation}
\text{FWHM}_{\text{image}} = 2\sqrt{\ln 2 \cdot (a^2 + b^2)}
\end{equation}

\begin{equation}
\text{CLASS\_STAR} = \frac{1}{1 + \exp\!\bigl(3 \cdot (r_{\text{FWHM}} - 1.5)\bigr)}, \qquad
r_{\text{FWHM}} = \frac{\text{FWHM}_{\text{image}}}{\text{FWHM}_{\text{seeing}} / s}
\end{equation}

where $s$ is the pixel scale in arcsec/pixel and $\text{FWHM}_{\text{seeing}}$ is set by \texttt{--seeing-fwhm} (default: 3.0 arcsec).

Objects with $\text{CLASS\_STAR} \approx 1$ are point-like (stars), while $\text{CLASS\_STAR} \approx 0$ indicates extended sources (galaxies).

\begin{remark}[Classifier Limitations]
This is a simple heuristic classifier, not the neural-network-based classifier used by SExtractor. For accurate star/galaxy separation, use dedicated tools such as \texttt{PSFEx} + SExtractor or machine learning classifiers.
\end{remark}


\section{Flags}

The \texttt{FLAGS} column is a bitmask combining detection and photometry flags:

\begin{center}
\begin{tabular}{ccl}
\toprule
\textbf{Bit} & \textbf{Value} & \textbf{Meaning} \\
\midrule
0 & 1 & Object has neighbors (bright and close enough to bias photometry) \\
1 & 2 & Object was originally blended with another \\
2 & 4 & At least one pixel is saturated (peak $+$ background $\geq$ \texttt{SATURATE}) \\
3 & 8 & Object is truncated (too close to image boundary) \\
4 & 16 & Aperture data incomplete or corrupted \\
5 & 32 & Isophotal data incomplete or corrupted \\
\bottomrule
\end{tabular}
\end{center}

Multiple flags can be set simultaneously. For example, \texttt{FLAGS = 3} means the object has neighbors (1) and was deblended (2).

\section{WCS Coordinate Transformation}

If the FITS header contains valid World Coordinate System (WCS) keywords (\texttt{CRPIX}, \texttt{CRVAL}, \texttt{CD} or \texttt{CDELT}+\texttt{CROTA}), pixel coordinates are converted to celestial coordinates (J2000 equatorial):

\begin{codebox}[WCS transformation]
\begin{lstlisting}[style=python,numbers=none]
wcs = WCS(header)
coords = wcs.pixel_to_world(objects['x'], objects['y'])
alpha_j2000 = coords.ra.deg    # Right Ascension in degrees
delta_j2000 = coords.dec.deg   # Declination in degrees
\end{lstlisting}
\end{codebox}

If WCS is unavailable, \texttt{ALPHA\_J2000} and \texttt{DELTA\_J2000} are set to 0.0.


%======================================================
\chapter{Output Parameters}
\label{ch:parameters}
%======================================================

This chapter provides a detailed description of each column in the output catalog.

\section{Complete Parameter Reference}

\begin{center}
\begin{longtable}{L{3.2cm} C{1.8cm} L{7.5cm}}
\toprule
\textbf{Parameter} & \textbf{Unit} & \textbf{Description} \\
\midrule
\endhead

\texttt{NUMBER} & --- & Running object number, assigned after sorting by brightness (brightest = 1). \\
\midrule

\texttt{X\_IMAGE} & pixel & Object centroid $x$-coordinate in FITS convention (1-indexed). Computed as the intensity-weighted first moment. \\
\midrule

\texttt{Y\_IMAGE} & pixel & Object centroid $y$-coordinate in FITS convention (1-indexed). \\
\midrule

\texttt{ALPHA\_J2000} & deg & Right Ascension (J2000 equatorial) computed from pixel coordinates via WCS. Zero if no WCS is available. \\
\midrule

\texttt{DELTA\_J2000} & deg & Declination (J2000 equatorial) computed from pixel coordinates via WCS. Zero if no WCS is available. \\
\midrule

\texttt{MAG\_AUTO} & mag & Kron-like elliptical aperture magnitude. Uses an elliptical aperture with semi-major axis $= 2.5 \times r_{\text{Kron}}$. Most reliable total magnitude for galaxies. \\
\midrule

\texttt{MAG\_ISOCOR} & mag & Isophotal magnitude corrected for flux lost below the detection threshold, using the SExtractor analytical correction formula. \\
\midrule

\texttt{MAG\_APER} & mag & Fixed circular aperture magnitude. Aperture diameter set by \texttt{--phot-aperture} (default: 5.0 pixels). Useful for comparing point sources. \\
\midrule

\texttt{FLUX\_AUTO} & counts & Total flux within the Kron elliptical aperture (in ADU or counts). \\
\midrule

\texttt{FLUX\_RADIUS} & pixel & Half-light radius: radius of a circular aperture enclosing 50\% of \texttt{FLUX\_AUTO}. Indicator of object size; correlates with effective radius $R_e$ for galaxies. \\
\midrule

\texttt{A\_IMAGE} & pixel & Semi-major axis of the object ellipse, derived from second-order intensity moments. \\
\midrule

\texttt{B\_IMAGE} & pixel & Semi-minor axis of the object ellipse. \\
\midrule

\texttt{THETA\_IMAGE} & deg & Position angle of the major axis, measured counter-clockwise from the $x$-axis (FITS convention). Range: $-90\degree$ to $+90\degree$. \\
\midrule

\texttt{ELLIPTICITY} & --- & Ellipticity: $e = 1 - b/a$. Values range from 0 (circular) to 1 (infinitely elongated). \\
\midrule

\texttt{KRON\_RADIUS} & pixel & First-moment radius (Kron radius), used to scale the AUTO aperture. Minimum enforced value: 3.5 pixels. \\
\midrule

\texttt{MU\_MAX} & mag/arcsec$^2$ & Peak surface brightness: magnitude of the brightest pixel divided by pixel area. Lower values = brighter peaks. \\
\midrule

\texttt{MU\_THRESHOLD} & mag/arcsec$^2$ & Detection threshold surface brightness at the object position. Objects with $\mu_{\max} < \mu_{\text{thresh}}$ are detected. \\
\midrule

\texttt{CLASS\_STAR} & --- & Star/galaxy classifier output. Range 0.0 (extended/galaxy) to 1.0 (point-like/star). Based on FWHM ratio to seeing. \\
\midrule

\texttt{FLAGS} & --- & Extraction flags (bitmask). See Section~\ref{ch:algorithm} for flag definitions. Zero means clean photometry. \\

\bottomrule
\end{longtable}
\end{center}


%======================================================
\chapter{Command-Line Reference}
\label{ch:cmdline}
%======================================================

The \sextract{} tool can also be used as a standalone command-line program, independent of the DS9 GUI.

\section{Synopsis}

\begin{codebox}
\begin{lstlisting}[style=bash,numbers=none]
ds9_sextract [OPTIONS] <FITS_FILE>
\end{lstlisting}
\end{codebox}

The output is a tab-separated table printed to \texttt{stdout}. The first line is the header row containing column names; subsequent lines contain data rows sorted by \texttt{MAG\_AUTO} (brightest first).

\section{Options}

\begin{center}
\begin{longtable}{L{4.0cm} C{1.5cm} L{7.0cm}}
\toprule
\textbf{Option} & \textbf{Default} & \textbf{Description} \\
\midrule
\endhead

\texttt{--detect-thresh} & 1.5 & Detection threshold in units of the local background RMS ($\sigma$). Lower values detect fainter sources but increase false positives. \\
\midrule

\texttt{--detect-minarea} & 5 & Minimum number of connected pixels above the threshold required to register a detection. Rejects cosmic ray hits and noise spikes. \\
\midrule

\texttt{--deblend-nthresh} & 32 & Number of exponentially spaced sub-thresholds used for deblending overlapping sources. Higher values provide finer deblending. \\
\midrule

\texttt{--deblend-mincont} & 0.005 & Minimum contrast ratio for deblending. A sub-component must contain at least this fraction of the parent flux to be split off. Lower values split more aggressively. \\
\midrule

\texttt{--phot-aperture} & 5.0 & Diameter of the fixed circular aperture (in pixels) used for \texttt{MAG\_APER} measurements. \\
\midrule

\texttt{--mag-zeropoint} & 25.0 & Photometric zero-point $m_0$ for magnitude conversion: $m = -2.5\log_{10}(F) + m_0$. Set this to the calibrated zero-point of your image. \\
\midrule

\texttt{--gain} & 0 & Detector gain in $e^-$/ADU. Used for Poisson noise estimation in photometric errors. If 0, the tool attempts to read \texttt{GAIN}, \texttt{EGAIN}, or \texttt{CCDGAIN} from the FITS header. \\
\midrule

\texttt{--pixel-scale} & 1.0 & Pixel scale in arcsec/pixel. Used for surface brightness calculations. If 1.0 and WCS is available, the scale is automatically derived from WCS. \\
\midrule

\texttt{--seeing-fwhm} & 3.0 & Seeing FWHM in arcsec. Used for star/galaxy classification (\texttt{CLASS\_STAR}). \\
\midrule

\texttt{--back-size} & 64 & Background mesh size in pixels. Each mesh cell has dimensions \texttt{back-size} $\times$ \texttt{back-size}. Smaller values track finer background variations but may subtract source flux. \\
\midrule

\texttt{--back-filtersize} & 3 & Median filter kernel size (in mesh cells) applied to the background map before interpolation. Smooths out mesh-to-mesh noise. \\

\bottomrule
\end{longtable}
\end{center}


\section{Usage Examples}

\begin{codebox}[Basic extraction]
\begin{lstlisting}[style=bash,numbers=none]
ds9_sextract image.fits > catalog.tsv
\end{lstlisting}
\end{codebox}

\begin{codebox}[Deep extraction with calibrated zero-point]
\begin{lstlisting}[style=bash,numbers=none]
ds9_sextract --detect-thresh 1.0 --detect-minarea 3 \
             --mag-zeropoint 28.09 --gain 2.5 \
             --seeing-fwhm 0.8 --pixel-scale 0.03 \
             hst_acs_f814w.fits > deep_catalog.tsv
\end{lstlisting}
\end{codebox}

\begin{codebox}[Conservative extraction (few false positives)]
\begin{lstlisting}[style=bash,numbers=none]
ds9_sextract --detect-thresh 3.0 --detect-minarea 10 \
             --deblend-mincont 0.01 \
             crowded_field.fits > clean_catalog.tsv
\end{lstlisting}
\end{codebox}

\begin{codebox}[Fine background estimation for extended objects]
\begin{lstlisting}[style=bash,numbers=none]
ds9_sextract --back-size 128 --back-filtersize 5 \
             galaxy_cluster.fits > cluster_catalog.tsv
\end{lstlisting}
\end{codebox}


%------------------------------------------------------
% PART II
%------------------------------------------------------
\part{Developer Reference}

%======================================================
\chapter{Tcl/Tk Implementation Details}
\label{ch:tcl}
%======================================================

This chapter documents the Tcl procedures that implement the catalog panel and its integration with the DS9 GUI.

\section{Catalog Panel Procedures}

\subsection{CreateCatalogPanel}

Creates all widgets for the catalog panel within \texttt{ds9(catalog\_frame)}:

\begin{center}
\begin{longtable}{L{3.5cm} L{9.5cm}}
\toprule
\textbf{Widget} & \textbf{Description} \\
\midrule
\endhead
\texttt{titlebar} & Horizontal frame containing the ``Source Extractor'' label, ``Extract'' button, and ``Clear'' button. \\
\midrule
\texttt{searchbar} & Filter entry with ``Apply'' button. The entry is bound to \texttt{<Return>} for keyboard activation. \\
\midrule
\texttt{tblf} (table frame) & Contains the \texttt{Tktable} widget (\texttt{catpanel(tbl)}) with horizontal and vertical scrollbars. The table is configured with 19 columns, row selection mode, and resizable column borders. \\
\midrule
\texttt{statusbar} & Displays status messages via the \texttt{catpanel(status)} variable. \\
\bottomrule
\end{longtable}
\end{center}

\subsection{CatalogPanelExtract}

Orchestrates the extraction process:

\begin{enumerate}[leftmargin=2em]
  \item Locates the \sextract{} binary in the same directory as the DS9 executable
  \item Retrieves the current FITS filename from the active frame
  \item Constructs \texttt{LD\_LIBRARY\_PATH} to include conda libraries (both \texttt{\$CONDA\_PREFIX/lib} and \texttt{\$HOME/miniconda3/lib})
  \item Executes \sextract{} via \texttt{exec sh -c} and captures stdout
  \item Passes the TSV output to \texttt{CatalogPanelLoadTSV}
\end{enumerate}

\subsection{CatalogPanelLoadTSV}

Parses tab-separated data into the Tktable widget:

\begin{enumerate}[leftmargin=2em]
  \item Splits the input string by newlines into rows
  \item Parses the first row as column headers
  \item Configures the table dimensions (\texttt{-cols}, \texttt{-rows})
  \item Fills the \texttt{catpaneltbldb} array variable: \texttt{catpaneltbldb(\textit{row},\textit{col})}
  \item Stores the raw data in \texttt{catpanel(alldata)} for filtering
\end{enumerate}

\subsection{CatalogPanelFilter}

Performs case-insensitive substring filtering:

\begin{enumerate}[leftmargin=2em]
  \item Reads the filter pattern from \texttt{catpanel(search\_var)}
  \item Iterates over stored data rows (\texttt{catpanel(alldata)})
  \item Includes a row if the pattern appears anywhere in the tab-separated line (\texttt{string match -nocase})
  \item Repopulates the table with matching rows only
  \item Updates the status bar with the count of matching rows
\end{enumerate}

\subsection{CatalogPanelAutoExtract}

A hook procedure called from \texttt{fits.tcl} after a FITS file is loaded:

\begin{codebox}
\begin{lstlisting}[style=tcl,numbers=none]
proc CatalogPanelAutoExtract {} {
    global catpanel
    if {[info exists catpanel(tbl)]} {
        after 500 CatalogPanelExtract
    }
}
\end{lstlisting}
\end{codebox}

The 500\,ms delay (\texttt{after 500}) ensures the image is fully loaded and rendered before extraction begins.


\section{Global Variables}

\begin{center}
\begin{longtable}{L{4.0cm} L{9.0cm}}
\toprule
\textbf{Variable} & \textbf{Purpose} \\
\midrule
\endhead
\texttt{ds9(toppw)} & The top-level \texttt{ttk::panedwindow} widget path \\
\texttt{ds9(catalog\_frame)} & Frame widget path for the right (catalog) pane \\
\texttt{catpanel(tbl)} & The Tktable widget path \\
\texttt{catpanel(tbldb)} & Name of the global array backing the table data \\
\texttt{catpanel(status)} & Status bar text variable \\
\texttt{catpanel(search\_var)} & Filter entry text variable \\
\texttt{catpanel(alldata)} & Raw TSV string from last extraction (for filtering) \\
\texttt{catpanel(filename)} & Last extracted FITS filename \\
\texttt{catpanel(delim)} & Column delimiter (always ``\texttt{\textbackslash t}'') \\
\bottomrule
\end{longtable}
\end{center}


%======================================================
\chapter{Python Extractor Details}
\label{ch:python}
%======================================================

\section{Module Structure}

The \sextract{} script (\texttt{ds9\_sextract.py}) is a single-file Python module with one main function:

\begin{codebox}
\begin{lstlisting}[style=python,numbers=none]
def extract_sources(fits_file,
                    detect_thresh=1.5,
                    detect_minarea=5,
                    deblend_nthresh=32,
                    deblend_mincont=0.005,
                    phot_aperture=5.0,
                    mag_zeropoint=25.0,
                    gain=0.0,
                    pixel_scale=1.0,
                    seeing_fwhm=3.0,
                    back_size=64,
                    back_filtersize=3):
\end{lstlisting}
\end{codebox}

\section{FITS File Handling}

The FITS reader searches for a 2D image HDU in the following order:

\begin{enumerate}[leftmargin=2em]
  \item Iterate through all HDUs; use the first one with 2D data (\texttt{hdu.data.ndim == 2})
  \item If no 2D HDU is found, try the primary HDU
  \item If the primary has $>2$ dimensions, extract the first 2D slice (\texttt{data[0]})
  \item The data is converted to \texttt{float64} C-contiguous format (required by SEP)
\end{enumerate}

\section{Automatic Header Parsing}

Several parameters are automatically read from the FITS header when not specified by the user:

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Parameter} & \textbf{Header Keywords} & \textbf{Fallback} \\
\midrule
Gain & \texttt{GAIN}, \texttt{EGAIN}, \texttt{CCDGAIN} & 0 (Poisson noise disabled) \\
Saturation level & \texttt{SATURATE}, \texttt{SATUR}, \texttt{SATULEVEL}, \texttt{DATAMAX} & 65535 \\
Pixel scale & WCS \texttt{proj\_plane\_pixel\_scales()} & 1.0 arcsec/pixel \\
\bottomrule
\end{tabular}
\end{center}

\section{Dependencies}

\begin{center}
\begin{longtable}{L{3cm} L{2.5cm} L{7cm}}
\toprule
\textbf{Package} & \textbf{Version} & \textbf{Purpose} \\
\midrule
\endhead
\texttt{sep} & $\geq 1.0$ & Core source extraction: background estimation, detection, deblending, photometry, and morphology. Python binding of the C library derived from SExtractor. \\
\midrule
\texttt{astropy} & $\geq 4.0$ & FITS file I/O (\texttt{astropy.io.fits}) and WCS coordinate transformations (\texttt{astropy.wcs.WCS}). \\
\midrule
\texttt{numpy} & $\geq 1.20$ & Array operations, mathematical functions, and data type management. \\
\bottomrule
\end{longtable}
\end{center}


%======================================================
\chapter{Image--Catalog Interaction}
\label{ch:interaction}
%======================================================

This chapter describes the interactive features that link the image display with the catalog panel, enabling bidirectional navigation, source merging, and advanced filtering.

\section{Overview of Interaction Features}

The following features connect the image markers to the catalog table:

\begin{center}
\begin{longtable}{L{3.5cm} L{9.5cm}}
\toprule
\textbf{Feature} & \textbf{Description} \\
\midrule
\endhead
Mark All & Places yellow ellipse markers on all detected sources in the image \\
\midrule
Marker Click & Clicking a yellow marker in the image selects and scrolls to the corresponding row in the catalog table, and shows cyan cross + green ellipse markers at the source position \\
\midrule
Visible Filter & Toggles display to show only sources within the current viewport \\
\midrule
Ctrl+Click Merge & Ctrl+Click on markers to select merge candidates (red thick ellipses); Ctrl+M merges them into a single source \\
\midrule
Trim & Column-based min/max filtering with a dialog for each catalog column \\
\midrule
Settings & Adjustable extraction parameters (threshold, deblending, photometry, etc.) \\
\bottomrule
\end{longtable}
\end{center}

\subsection{Title Bar Layout}

\begin{codebox}
\begin{lstlisting}[numbers=none]
[Source Extractor]  [Extract] [Settings...] [Mark All] [Visible] [Trim...] [Clear]
\end{lstlisting}
\end{codebox}


\section{Mark All Sources}

The \textbf{Mark All} button creates yellow ellipse markers on all detected sources using the \texttt{ISO\_RADIUS}, \texttt{B\_IMAGE}/\texttt{A\_IMAGE} ratio, and \texttt{THETA\_IMAGE} parameters:

\begin{codebox}[Marker region string]
\begin{lstlisting}[style=tcl,numbers=none]
ellipse($x $y ${semi_a}i ${semi_b}i $theta)
  # color=yellow width=1 tag={sextract_all} tag={sextract_src.$num}
    select=0 edit=0 move=0 rotate=0 delete=1 highlite=1
    callback=highlite CatalogPanelMarkerCB {$num}
    callback=unhighlite CatalogPanelMarkerUnCB {$num}
\end{lstlisting}
\end{codebox}

Each marker has:
\begin{itemize}[leftmargin=2em]
  \item \texttt{sextract\_all} tag for bulk operations (deletion, etc.)
  \item \texttt{sextract\_src.\textit{NUMBER}} tag for individual identification
  \item \texttt{highlite=1} with callbacks for mouse interaction in \texttt{catalog} mode
\end{itemize}


\section{Marker Click --- Image to Catalog Navigation}

Clicking a yellow ellipse marker in the image navigates the catalog to the corresponding source.

\subsection{Implementation}

DS9's built-in \texttt{highlite} callbacks only fire in \texttt{catalog} mode. Since the default viewing mode is \texttt{none}, explicit click handlers were added in \texttt{frame.tcl}:

\begin{codebox}[frame.tcl --- Button1Frame (none mode)]
\begin{lstlisting}[style=tcl,numbers=none]
none {
    if {$which == $current(frame)} {
        CatalogPanelMarkerClick $which $x $y
    } else {
        ...
    }
}
\end{lstlisting}
\end{codebox}

The click detection uses DS9's marker API:

\begin{codebox}[CatalogPanelMarkerClick]
\begin{lstlisting}[style=tcl,numbers=none]
proc CatalogPanelMarkerClick {which x y} {
    set id [$which get marker catalog id $x $y]
    if {$id == 0} return
    set tags [$which get marker catalog $id tag]
    foreach tag $tags {
        if {[string match "sextract_src.*" $tag]} {
            set src_num [string range $tag 13 end]
            break
        }
    }
    CatalogPanelMarkerCB $src_num $id
}
\end{lstlisting}
\end{codebox}

\texttt{CatalogPanelMarkerCB} then:
\begin{enumerate}[leftmargin=2em]
  \item Finds the \texttt{NUMBER} column in the table header
  \item Searches all rows for the matching source number
  \item Selects the row and scrolls the table to it
  \item Calls \texttt{CatalogPanelGotoSource} to pan the image and show selection markers (cyan cross + green ellipse)
\end{enumerate}


\section{Visible Filter}

The \textbf{Visible} button toggles between showing all sources and showing only sources within the current viewport.

\subsection{Viewport Calculation}

\begin{codebox}[Viewport bounds computation]
\begin{lstlisting}[style=tcl,numbers=none]
set center [$frame get cursor image]   ;# {cx cy}
set zoom   [$frame get zoom]           ;# {zx zy}
set cw [winfo width $ds9(canvas)]
set ch [winfo height $ds9(canvas)]
set x_min [expr {$cx - $cw / 2.0 / $zx}]
set x_max [expr {$cx + $cw / 2.0 / $zx}]
set y_min [expr {$cy - $ch / 2.0 / $zy}]
set y_max [expr {$cy + $ch / 2.0 / $zy}]
\end{lstlisting}
\end{codebox}

Sources with \texttt{X\_IMAGE} and \texttt{Y\_IMAGE} within these bounds are displayed. The status bar shows ``Visible: N of M sources in current view.''


\section{Source Merging (Ctrl+Click + Ctrl+M)}

Multiple sources can be merged into a single source through a two-step process.

\subsection{Step 1: Select Merge Candidates}

\textbf{Ctrl+Click} on a yellow ellipse marker toggles it as a merge candidate:
\begin{itemize}[leftmargin=2em]
  \item Selected: A red thick ellipse (\texttt{width=3}) is overlaid, tagged \texttt{sextract\_merge} and \texttt{sextract\_merge.\textit{NUMBER}}
  \item Deselected: Ctrl+Click again removes the red marker and the source from the merge list
  \item Status bar: ``Merge: N sources selected (Ctrl+M to merge, Esc to cancel)''
\end{itemize}

\subsection{Step 2: Execute Merge (Ctrl+M)}

When at least 2 sources are selected, \textbf{Ctrl+M} merges them:

\begin{center}
\begin{longtable}{L{3.2cm} L{9.3cm}}
\toprule
\textbf{Property} & \textbf{Merge Algorithm} \\
\midrule
\endhead
\texttt{X\_IMAGE}, \texttt{Y\_IMAGE} & Flux-weighted centroid: $\bar{x} = \sum F_i x_i / \sum F_i$ \\
\midrule
\texttt{FLUX\_AUTO} & Sum of all fluxes: $F_{\text{total}} = \sum F_i$ \\
\midrule
\texttt{MAG\_AUTO} & Recomputed: $m = -2.5\log_{10}(F_{\text{total}}) + m_0$ \\
\midrule
\texttt{NPIX\_ISO} & Sum of all pixel counts \\
\midrule
\texttt{ISO\_RADIUS} & $r = \sqrt{N_{\text{pix,total}} / (\pi \cdot b/a)}$ \\
\midrule
\texttt{A\_IMAGE}, \texttt{B\_IMAGE}, \texttt{THETA\_IMAGE} & Flux-weighted average \\
\midrule
\texttt{NUMBER} & New number: $\max(\text{all NUMBERs}) + 1$ \\
\midrule
Other columns & Copied from the brightest source \\
\bottomrule
\end{longtable}
\end{center}

After merging:
\begin{enumerate}[leftmargin=2em]
  \item The original sources are removed from \texttt{catpanel(alldata)}
  \item The merged source row is appended
  \item The table and markers are regenerated
  \item The merged source is auto-selected and the image pans to its position
  \item Status bar: ``Merged N sources into \#NEW: pos=(X,Y) mag=M''
\end{enumerate}

\subsection{Cancel (Escape)}

Pressing \textbf{Escape} cancels the merge selection, removes all red markers, and clears the merge list.


\section{Trim --- Column Filter Dialog}

The \textbf{Trim...} button opens a modal dialog for per-column min/max filtering.

\subsection{Dialog Layout}

\begin{codebox}[Trim dialog structure]
\begin{lstlisting}[numbers=none]
+------------------------------------------+
|  Trim / Column Filter                    |
+------------------------------------------+
|  Column       |  Min      |  Max         |
|  NUMBER       | [______]  | [______]     |
|  X_IMAGE      | [______]  | [______]     |
|  Y_IMAGE      | [______]  | [______]     |
|  ...          | ...       | ...          |
|  MAG_AUTO     | [______]  | [______]     |
+------------------------------------------+
|  [Apply] [Reset] [Save] [Load] [Cancel]  |
+------------------------------------------+
\end{lstlisting}
\end{codebox}

\begin{itemize}[leftmargin=2em]
  \item Empty fields mean no constraint
  \item \textbf{Apply}: filter rows where $\text{min} \leq \text{value} \leq \text{max}$ for each specified column, then close dialog and regenerate markers
  \item \textbf{Reset}: clear all min/max fields
  \item \textbf{Save/Load}: persist conditions to \texttt{\textasciitilde/.ds9/sextract\_trim.prf}
  \item \textbf{Cancel}: close without changes
\end{itemize}

Status bar after Apply: ``Trimmed: N of M sources match conditions.''


\section{Settings Dialog}

The \textbf{Settings...} button opens a modal dialog to adjust extraction parameters before running Extract.

\begin{center}
\begin{longtable}{L{4.0cm} C{1.8cm} L{6.5cm}}
\toprule
\textbf{Parameter} & \textbf{Default} & \textbf{Description} \\
\midrule
\endhead
Detect Threshold & 1.5 & Detection threshold ($\sigma$) \\
Detect Min Area & 5 & Minimum connected pixels \\
Deblend NThresh & 32 & Number of deblending sub-thresholds \\
Deblend MinCont & 0.005 & Minimum deblending contrast \\
Phot Aperture & 5.0 & Fixed aperture diameter (pixels) \\
Mag Zeropoint & 25.0 & Magnitude zero-point $m_0$ \\
Gain & 0.0 & Detector gain ($e^-$/ADU) \\
Pixel Scale & 1.0 & Pixel scale (arcsec/pixel) \\
Seeing FWHM & 3.0 & Seeing FWHM (arcsec) \\
Back Size & 64 & Background mesh size (pixels) \\
Back Filter Size & 3 & Background filter kernel (mesh cells) \\
\bottomrule
\end{longtable}
\end{center}

Settings are saved to / loaded from \texttt{\textasciitilde/.ds9/sextract.prf}. The \textbf{Defaults} button restores all parameters to their default values.


\section{Marker Tag Reference}

\begin{center}
\begin{longtable}{L{4.2cm} L{8.8cm}}
\toprule
\textbf{Tag} & \textbf{Description} \\
\midrule
\endhead
\texttt{sextract\_all} & All yellow ellipse markers from Mark All \\
\midrule
\texttt{sextract\_src.\textit{N}} & Individual source marker (N = source NUMBER) \\
\midrule
\texttt{sextract\_sel} & Current selection markers (cyan cross + green ellipse) \\
\midrule
\texttt{sextract\_merge} & All red merge-candidate markers \\
\midrule
\texttt{sextract\_merge.\textit{N}} & Individual merge-candidate marker \\
\bottomrule
\end{longtable}
\end{center}


\section{Key Bindings}

\begin{center}
\begin{tabular}{ll}
\toprule
\textbf{Key} & \textbf{Action} \\
\midrule
Click on marker & Navigate catalog to source, show selection markers \\
Ctrl+Click on marker & Toggle merge candidate (red ellipse) \\
Ctrl+M & Merge selected sources \\
Escape & Cancel merge selection \\
Click table header & Sort by column (toggle ascending/descending) \\
Click table row & Pan image to source position \\
\bottomrule
\end{tabular}
\end{center}


%======================================================
\chapter{Building \sextract{} for Each Platform}
\label{ch:building}
%======================================================

The \sextract{} binary is a standalone C program that uses the SEP library and CFITSIO for source extraction. It must be compiled separately for each target platform and placed in the \texttt{bin/} directory alongside the \ds{} executable.

\section{Dependencies}

\begin{center}
\begin{tabular}{lll}
\toprule
\textbf{Library} & \textbf{Purpose} & \textbf{Install} \\
\midrule
CFITSIO & FITS file I/O & Package manager or source \\
SEP & Source extraction algorithm & Included in \texttt{ds9\_sextract.c} or separate \\
libm & Math functions & System standard \\
\bottomrule
\end{tabular}
\end{center}


\section{Linux (x86\_64)}

\begin{codebox}[Install dependencies (Ubuntu/Debian)]
\begin{lstlisting}[style=bash,numbers=none]
sudo apt-get install libcfitsio-dev
\end{lstlisting}
\end{codebox}

\begin{codebox}[Install dependencies (RHEL/CentOS/Rocky)]
\begin{lstlisting}[style=bash,numbers=none]
sudo dnf install cfitsio-devel
\end{lstlisting}
\end{codebox}

\begin{codebox}[Install dependencies (conda)]
\begin{lstlisting}[style=bash,numbers=none]
conda install -c conda-forge cfitsio
\end{lstlisting}
\end{codebox}

\begin{codebox}[Compile]
\begin{lstlisting}[style=bash,numbers=none]
gcc -O2 -o ds9_sextract ds9_sextract.c \
    -I/usr/include -L/usr/lib64 \
    -lcfitsio -lm
# Or with conda:
gcc -O2 -o ds9_sextract ds9_sextract.c \
    -I$CONDA_PREFIX/include -L$CONDA_PREFIX/lib \
    -lcfitsio -lm -Wl,-rpath,$CONDA_PREFIX/lib
cp ds9_sextract /path/to/SAOImageDS9/bin/
\end{lstlisting}
\end{codebox}


\section{macOS (Intel and Apple Silicon)}

\begin{codebox}[Install dependencies (Homebrew)]
\begin{lstlisting}[style=bash,numbers=none]
brew install cfitsio
\end{lstlisting}
\end{codebox}

\begin{codebox}[Install dependencies (conda)]
\begin{lstlisting}[style=bash,numbers=none]
conda install -c conda-forge cfitsio
\end{lstlisting}
\end{codebox}

\begin{codebox}[Compile (Homebrew)]
\begin{lstlisting}[style=bash,numbers=none]
# Intel Mac:
gcc -O2 -o ds9_sextract ds9_sextract.c \
    -I/usr/local/include -L/usr/local/lib \
    -lcfitsio -lm

# Apple Silicon (M1/M2/M3):
gcc -O2 -o ds9_sextract ds9_sextract.c \
    -I/opt/homebrew/include -L/opt/homebrew/lib \
    -lcfitsio -lm
\end{lstlisting}
\end{codebox}

\begin{codebox}[Compile (conda)]
\begin{lstlisting}[style=bash,numbers=none]
gcc -O2 -o ds9_sextract ds9_sextract.c \
    -I$CONDA_PREFIX/include -L$CONDA_PREFIX/lib \
    -lcfitsio -lm -Wl,-rpath,$CONDA_PREFIX/lib
\end{lstlisting}
\end{codebox}

\begin{codebox}[Place binary]
\begin{lstlisting}[style=bash,numbers=none]
cp ds9_sextract /path/to/SAOImageDS9/bin/
\end{lstlisting}
\end{codebox}

\begin{important}
On macOS, if you encounter ``library not loaded'' errors at runtime, ensure the CFITSIO library path is embedded using \texttt{-Wl,-rpath} or set \texttt{DYLD\_LIBRARY\_PATH} before launching DS9.
\end{important}


\section{Windows}

\subsection{Using MSYS2/MinGW-w64}

\begin{codebox}[Install MSYS2 and dependencies]
\begin{lstlisting}[style=bash,numbers=none]
# In MSYS2 MinGW64 shell:
pacman -S mingw-w64-x86_64-gcc mingw-w64-x86_64-cfitsio
\end{lstlisting}
\end{codebox}

\begin{codebox}[Compile]
\begin{lstlisting}[style=bash,numbers=none]
gcc -O2 -o ds9_sextract.exe ds9_sextract.c \
    -I/mingw64/include -L/mingw64/lib \
    -lcfitsio -lm
\end{lstlisting}
\end{codebox}

\begin{codebox}[Deploy]
\begin{lstlisting}[style=bash,numbers=none]
# Copy binary and required DLLs to ds9 bin directory
cp ds9_sextract.exe /path/to/SAOImageDS9/bin/
cp /mingw64/bin/libcfitsio*.dll /path/to/SAOImageDS9/bin/
cp /mingw64/bin/zlib1.dll /path/to/SAOImageDS9/bin/
\end{lstlisting}
\end{codebox}

\subsection{Using Visual Studio}

\begin{codebox}[Compile with MSVC]
\begin{lstlisting}[style=bash,numbers=none]
cl /O2 /Fe:ds9_sextract.exe ds9_sextract.c ^
   /I"C:\cfitsio\include" ^
   /link /LIBPATH:"C:\cfitsio\lib" cfitsio.lib
\end{lstlisting}
\end{codebox}

\begin{important}
On Windows, required DLLs (\texttt{libcfitsio*.dll}, \texttt{zlib1.dll}) must be placed in the same directory as \texttt{ds9\_sextract.exe}, or their location must be in the system \texttt{PATH}.
\end{important}


\section{Cross-Platform Tcl Integration}

The \texttt{CatalogPanelExtract} procedure in \texttt{layout.tcl} automatically handles platform differences:

\begin{codebox}[Platform-aware binary discovery]
\begin{lstlisting}[style=tcl,numbers=none]
set os $::tcl_platform(os)
if {$os eq "Windows NT"} {
    set sextract [file join $bindir ds9_sextract.exe]
} else {
    set sextract [file join $bindir ds9_sextract]
}
\end{lstlisting}
\end{codebox}

\begin{codebox}[Platform-aware library path]
\begin{lstlisting}[style=tcl,numbers=none]
if {$os eq "Darwin"} {
    # macOS: set DYLD_LIBRARY_PATH
    set ::env(DYLD_LIBRARY_PATH) [join $libpaths :]
} elseif {$os ne "Windows NT"} {
    # Linux/Unix: set LD_LIBRARY_PATH
    set ::env(LD_LIBRARY_PATH) [join $libpaths :]
}
# Windows: DLLs found via PATH automatically
\end{lstlisting}
\end{codebox}

\begin{codebox}[Cross-platform execution]
\begin{lstlisting}[style=tcl,numbers=none]
# Direct exec without sh -c (works on all platforms)
exec $sextract $fn {*}$paramargs 2>@stderr
\end{lstlisting}
\end{codebox}

\begin{center}
\begin{tabular}{lccc}
\toprule
 & \textbf{Linux} & \textbf{macOS} & \textbf{Windows} \\
\midrule
Binary name & \texttt{ds9\_sextract} & \texttt{ds9\_sextract} & \texttt{ds9\_sextract.exe} \\
Library path env & \texttt{LD\_LIBRARY\_PATH} & \texttt{DYLD\_LIBRARY\_PATH} & \texttt{PATH} \\
Shell required & No & No & No \\
\bottomrule
\end{tabular}
\end{center}


\section{Verification}

After building for any platform, verify the binary works:

\begin{codebox}[Test extraction]
\begin{lstlisting}[style=bash,numbers=none]
./ds9_sextract test_image.fits | head -5
\end{lstlisting}
\end{codebox}

Expected output: a tab-separated header line followed by data rows. If library errors occur, check the library path configuration for your platform.


%------------------------------------------------------
% Back matter
%------------------------------------------------------

\printbibliography[heading=bibintoc,title={References}]

\end{document}
